---
title: Analysis of weather events which most harmful with respect to population health
  and  have the greatest economic consequences across the United States.
author: "Nariman Daniyarov"
date: "25 12 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The basic goal of this project is to explore the NOAA Storm Database and answer next basic questions:  
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

## Data Processing
Include required libraries
```{r}
library(dplyr)
```


The data can be download as follow:
```{r}
data_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
data_file <- "repdata_data_StormData.csv.bz2"

if(!file.exists(data_file)) {
        download.file(data_url, data_file, mode = "wb")
}
```
Read the downloaded data:
```{r read-data, cache=TRUE}
raw_data <- read.csv(data_file, header = TRUE)
```
```{r}
data_dim <- dim(raw_data)
```
Data has **`r data_dim[1]`** observations and **`r data_dim[2]`** variables.  

For further work with data, we need to display a list of column names
```{r}
colnames(raw_data)
```

**#### Question 1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?**
For purpose of this, we have to subset the data selecting three columns
```{r}
hd_data <- select(raw_data, EVTYPE,  FATALITIES, INJURIES)
```
Check for NAs on FATALITIES and INJURIES columns
```{r}
sum(is.na(hd_data$FATALITIES))
sum(is.na(hd_data$INJURIES))
```
Group data by severe weather types, and summarise values on FATALITIES and INJURIES
```{r message=FALSE}
hd_data %>% 
        group_by(EVTYPE) %>% 
        summarise(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES)) %>% 
        arrange(desc(FATALITIES, INJURIES))
```




**#### Question 2. Across the United States, which types of events have the greatest economic consequences?**